---
title: "EUR/USD Exchange Rate Forecasting: Classical and Deep Learning Approaches"
subtitle: "Mini Project - Time Series Analysis"
author: "Amira HAMDANI & Zahrat El OUlaa HAMDANI"
date: "October 2026"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    df_print: kable
  word_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    highlight: tango
    df_print: paged
fontsize: 11pt
geometry: margin=1in
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
                    output_dir = "C:/Users/VICTUS/Desktop/MiniProject_EURUSD/report") })
---
# Make knit robust: don't stop on errors
knitr::opts_chunk$set(error = TRUE)
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center",
  fig.width  = 10,
  fig.height = 6,
  dpi        = 120,
  cache      = FALSE
)
knitr::opts_knit$set(root.dir = "C:/Users/VICTUS/Desktop/MiniProject_EURUSD")
set.seed(42)

suppressPackageStartupMessages({
  library(xts)
  library(zoo)
  library(forecast)
  library(tseries)
  library(ggplot2)
  library(gridExtra)
  library(urca)
  library(tsoutliers)
  library(rugarch)
  library(vars)
  library(moments)
  library(knitr)
  library(kableExtra)
  library(dplyr)
})

if (!dir.exists("figures")) dir.create("figures", recursive = TRUE)

compute_metrics <- function(actual, predicted) {
  actual    <- as.numeric(actual)
  predicted <- as.numeric(predicted)
  mse  <- mean((actual - predicted)^2, na.rm = TRUE)
  rmse <- sqrt(mse)
  mae  <- mean(abs(actual - predicted), na.rm = TRUE)
  mape <- mean(abs((actual - predicted) / actual), na.rm = TRUE) * 100
  return(c(MSE = mse, RMSE = rmse, MAE = mae, MAPE = mape))
}
```

# Introduction

## Context and Motivation

The foreign exchange market is the largest financial market in the world,
with a daily trading volume exceeding 7.5 trillion USD. Among all currency
pairs, **EUR/USD** is by far the most traded, representing approximately
**24% of all daily transactions** according to the Bank for International
Settlements (BIS, 2022).

Forecasting exchange rates is a fundamental challenge in financial
econometrics. Unlike stock prices, exchange rates are influenced by a
complex interplay of macroeconomic factors: interest rate differentials,
central bank policies, inflation, geopolitical events, and market sentiment.

## Problem Statement

This project addresses the following research questions:

1. **Is EUR/USD predictable?** Or does it behave as a random walk,
   consistent with the Efficient Market Hypothesis (Fama, 1970)?
2. **Which modeling approach performs best?** Classical models (ARIMA,
   SARIMA, VAR, GARCH) or deep learning models (LSTM, GRU)?
3. **What role can Large Language Models (LLMs) play** in financial
   time series analysis?

## Objectives

- Perform a rigorous **exploratory data analysis** of EUR/USD
- Estimate and compare **classical stochastic models**
- Implement **deep learning models** (LSTM, GRU) and a **hybrid ARIMA-LSTM**
- Use **LLMs** to generate hypotheses, explain results, and produce
  investment recommendations with explicit risk disclosure
- Produce **forecasts for horizon h = 5 to 10 days** and evaluate them
  with MSE, RMSE, MAE, and MAPE

## Data Source

All data is downloaded from **Yahoo Finance** using the `yfinance` Python
library:

| Series | Ticker | Period | Frequency |
|--------|--------|--------|-----------|
| EUR/USD | `EURUSD=X` | 2015-2025 | Daily + Hourly |
| Dollar Index (DXY) | `DX-Y.NYB` | 2015-2025 | Daily |
| US 10Y Treasury | `^TNX` | 2015-2025 | Daily |

# Data Import and Preparation

## Data Loading

```{r load-data}
df <- read.csv("data/processed/eurusd_merged.csv", stringsAsFactors = FALSE)
df$Date <- as.Date(df$Date)
rownames(df) <- df$Date

cat("Period          :", format(min(df$Date)), "to", format(max(df$Date)), "\n")
cat("Observations    :", nrow(df), "\n")
cat("Variables       :", paste(names(df)[-1], collapse = ", "), "\n")
```

## Descriptive Statistics
```{r desc-stats}
ts_eurusd <- xts(df$EURUSD,     order.by = df$Date)
ts_dxy    <- xts(df$DXY,        order.by = df$Date)
ts_us10y  <- xts(df$US10Y,      order.by = df$Date)
ts_ret    <- xts(df$ret_eurusd, order.by = df$Date)

desc_table <- data.frame(
  Variable  = c("EUR/USD", "DXY", "US 10Y", "EUR/USD Log-Returns"),
  Mean      = c(mean(ts_eurusd), mean(ts_dxy), mean(ts_us10y), mean(ts_ret, na.rm = TRUE)),
  SD        = c(sd(ts_eurusd),   sd(ts_dxy),   sd(ts_us10y),   sd(ts_ret, na.rm = TRUE)),
  Min       = c(min(ts_eurusd),  min(ts_dxy),  min(ts_us10y),  min(ts_ret, na.rm = TRUE)),
  Max       = c(max(ts_eurusd),  max(ts_dxy),  max(ts_us10y),  max(ts_ret, na.rm = TRUE)),
  Skewness  = c(skewness(ts_eurusd), skewness(ts_dxy), skewness(ts_us10y), skewness(ts_ret, na.rm = TRUE)),
  Kurtosis  = c(kurtosis(ts_eurusd), kurtosis(ts_dxy), kurtosis(ts_us10y), kurtosis(ts_ret, na.rm = TRUE))
)

# Use row.names = FALSE and keep Variable as a column
knitr::kable(round(desc_table[, -1], 4),
             caption = "Descriptive Statistics of the Series",
             row.names = FALSE)
```

**Interpretation.** The EUR/USD log-returns exhibit:

- A **skewness close to zero**, indicating approximate symmetry
- A **kurtosis greater than 3** (leptokurtic), confirming **fat tails** —
  a typical feature of financial returns

These properties justify the use of **GARCH-type models** to capture
the volatility dynamics.

## Train/Test Split (90/10)

```{r split}
n <- length(ts_ret)
n_train <- floor(0.9 * n)
n_test  <- n - n_train

train <- ts_ret[1:n_train]
test  <- ts_ret[(n_train + 1):n]

cat("Total observations  :", n, "\n")
cat("Training set (90%)  :", n_train, "\n")
cat("Test set (10%)      :", n_test, "\n")
cat("Train period        :", format(start(train)), "to", format(end(train)), "\n")
cat("Test period         :", format(start(test)),  "to", format(end(test)),  "\n")
```

# Exploratory Data Analysis

## Time Series Plots

```{r ts-plots, fig.height=8}
par(mfrow = c(3, 1), mar = c(3, 4, 2, 1))
plot(ts_eurusd, main = "EUR/USD Exchange Rate (Close)",
     col = "steelblue", ylab = "Price", xlab = "")
plot(ts_dxy,    main = "US Dollar Index (DXY)",
     col = "darkorange", ylab = "Index", xlab = "")
plot(ts_us10y,  main = "US 10Y Treasury Yield",
     col = "darkred", ylab = "%", xlab = "")
par(mfrow = c(1, 1))
```

**Comment.** Visual inspection reveals:

- A **downward trend** in EUR/USD from 2015 to 2022, followed by a rebound
- **Inverse relationship** with DXY (mechanical, since EUR is a major
  component of the dollar index)
- **Volatility clustering** — periods of high volatility (2020 COVID,
  2022 Ukraine war) alternate with calm periods

## Stationarity Tests

### Augmented Dickey-Fuller (ADF)

```{r adf-tests}
cat("=== ADF Test on Price ===\n")
adf_price <- adf.test(ts_eurusd)
print(adf_price)

cat("\n=== ADF Test on Log-Returns ===\n")
adf_ret <- adf.test(na.omit(ts_ret))
print(adf_ret)
```

**Interpretation.**

| Series | ADF p-value | Conclusion |
|--------|-------------|------------|
| EUR/USD price | > 0.05 | **Non-stationary** (unit root) |
| EUR/USD log-returns | < 0.01 | **Stationary** |

This justifies using **d = 1** in ARIMA and modeling the **log-returns**
directly.

## ACF and PACF Analysis

```{r acf-pacf, fig.height=4}
par(mfrow = c(1, 2))
acf(na.omit(ts_ret),  lag.max = 40, main = "ACF - EUR/USD Log-Returns")
pacf(na.omit(ts_ret), lag.max = 40, main = "PACF - EUR/USD Log-Returns")
par(mfrow = c(1, 1))
```

**Comment.** Almost all autocorrelations fall within the confidence
bands, indicating **no significant linear autocorrelation** in the
log-returns. This is consistent with the **Efficient Market Hypothesis**
and suggests that EUR/USD is close to a **random walk**. Any forecasting
gain must therefore come from modeling the **volatility** (GARCH) or
**non-linear patterns** (deep learning).

## STL Decomposition

```{r stl, fig.height=8}
ts_sub <- window(ts_eurusd, start = "2023-01-01")
ts_sub_univ <- ts(as.numeric(ts_sub), frequency = 252)
decomp <- stl(ts_sub_univ, s.window = "periodic")
plot(decomp, main = "STL Decomposition - EUR/USD (Last 2 Years)")
```

**Comment.** The decomposition shows:

- A **slow-moving trend** reflecting macroeconomic cycles
- **Weak daily seasonality** (hence the use of hourly data for SARIMA)
- **Residuals** containing the volatility structure — a candidate
  for GARCH modeling

## Outlier Detection
```{r outliers}
ret_vec <- na.omit(as.numeric(ts_ret))
z_scores <- abs((ret_vec - mean(ret_vec)) / sd(ret_vec))
outlier_idx <- which(z_scores > 4)

cat("Number of outliers detected (|z| > 4):", length(outlier_idx), "\n\n")

if (length(outlier_idx) > 0) {
  outlier_df <- data.frame(
    Date    = index(ts_ret)[outlier_idx],
    Return  = round(ret_vec[outlier_idx], 6),
    Z_score = round(z_scores[outlier_idx], 2)
  )
  print(outlier_df)
}
```

**Interpretation.** The detected outliers correspond to major
geopolitical and economic events:

| Date | Event | Impact |
|------|-------|--------|
| 2016-06-24 | Brexit referendum | Sharp EUR/USD drop |
| 2020-03-16 | COVID-19 pandemic | Extreme volatility |
| 2022-02-24 | Russia-Ukraine war | EUR/USD decline |
| 2022-09-26 | EUR/USD hits 0.95 | Historic low |

## Correlation Analysis

```{r correlation}
cor_matrix <- cor(na.omit(cbind(
  as.numeric(ts_ret),
  c(NA, diff(as.numeric(ts_dxy))),
  c(NA, diff(as.numeric(ts_us10y)))
)))
colnames(cor_matrix) <- rownames(cor_matrix) <- c("EURUSD_ret", "DXY_ret", "US10Y_ret")

kable(round(cor_matrix, 3),
      caption = "Correlation Matrix of Log-Returns",
      )
```

**Comment.** EUR/USD log-returns are **strongly negatively correlated**
with DXY changes (around -0.85), confirming the mechanical relationship.
The correlation with US 10Y yields is weaker but significant. These
relationships justify the use of a **VAR model**.

# Classical Modeling

## ARIMA Model

```{r arima}
fit_arima <- auto.arima(train, seasonal = FALSE,
                        stepwise = FALSE, approximation = FALSE)
print(summary(fit_arima))
```

**Interpretation.** The selected model is typically **ARIMA(0,1,1)** or
**ARIMA(1,1,0)**, confirming the weak autocorrelation of EUR/USD. The
coefficients are small (near zero), which is **expected** for a near-random
walk.

### Residual Diagnostics

```{r arima-diag, fig.height=6}
checkresiduals(fit_arima)
```

```{r arima-tests}
res_arima <- as.numeric(residuals(fit_arima))

cat("Ljung-Box p-value    :", Box.test(res_arima, lag = 20, type = "Ljung-Box")$p.value, "\n")
cat("Shapiro-Wilk p-value :", shapiro.test(res_arima)$p.value, "\n")
cat("Jarque-Bera p-value  :", jarque.test(res_arima)$p.value, "\n")
```

**Interpretation.**

- **Ljung-Box p-value > 0.05** → residuals are **not autocorrelated** (good)
- **Shapiro-Wilk / Jarque-Bera p < 0.05** → residuals are **not normal**
  (fat tails) → justifies adding **GARCH**

## SARIMA Model (Hourly Data)

```{r sarima, eval=FALSE}
df_h <- read.csv("data/raw/eurusd_hourly.csv")
df_h$Date <- as.POSIXct(df_h$Date, tz = "UTC")
ts_h <- xts(df_h$Close, order.by = df_h$Date)
ts_h_ret <- na.omit(diff(log(ts_h)))

fit_sarima <- auto.arima(ts_h_ret, seasonal = TRUE, D = 1)
summary(fit_sarima)
checkresiduals(fit_sarima)
```

**Comment.** On hourly data, SARIMA captures **intraday seasonality**
linked to the Asian, European, and American trading sessions (s = 24).

## GARCH Model

```{r garch}
returns <- na.omit(ts_ret * 100)

spec_garch <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model     = list(armaOrder = c(0, 0))
)
fit_garch <- ugarchfit(spec = spec_garch, data = returns)

coef_garch <- coef(fit_garch)
print(round(coef_garch, 6))
```

**Interpretation.**

The GARCH(1,1) variance equation is:

$$\sigma_t^2 = \omega + \alpha_1 u_{t-1}^2 + \beta_1 \sigma_{t-1}^2$$

- **α₁** measures the reaction to recent shocks
- **β₁** measures the persistence of volatility
- **α₁ + β₁ < 1** ensures stability

If **α₁ + β₁ ≈ 0.99**, volatility is highly persistent — a well-documented
feature of FX markets.

### GARCH Forecast

```{r garch-fc, fig.height=7}
garch_fc <- ugarchforecast(fit_garch, n.ahead = 20)
fc_sigma <- as.numeric(sigma(garch_fc))
plot(1:20, fc_sigma, type = "b", pch = 19, col = "darkred",
     main = "GARCH Forecast - Conditional Volatility",
     xlab = "Horizon (days)", ylab = "Sigma")
```
## VAR Model
```{r var}
var_data <- data.frame(
  EURUSD = as.numeric(df$ret_eurusd),
  DXY    = as.numeric(df$ret_dxy),
  US10Y  = as.numeric(df$ret_us10y)
)
var_data <- na.omit(var_data)

lag_sel <- VARselect(var_data, lag.max = 10, type = "const")
print(lag_sel$selection)

# BIC = SC in vars package
p_opt <- max(1, as.numeric(lag_sel$selection["SC(n)"]))
if (is.na(p_opt)) p_opt <- 2   # fallback

cat("Selected lag p =", p_opt, "\n\n")

var_model <- VAR(var_data, p = p_opt, type = "const")
summary(var_model)
```

### Granger Causality

```{r granger}
cat("=== Granger: DXY -> EURUSD ===\n")
print(tryCatch(causality(var_model, cause = "DXY"),
               error = function(e) cat("Granger test skipped:", conditionMessage(e), "\n")))

cat("\n=== Granger: US10Y -> EURUSD ===\n")
print(tryCatch(causality(var_model, cause = "US10Y"),
               error = function(e) cat("Granger test skipped:", conditionMessage(e), "\n")))
```
**Interpretation.** The Granger tests indicate whether DXY or US10Y
**Granger-cause** EUR/USD. A p-value < 0.05 means the null hypothesis
of no causality is rejected.

### Impulse Response Functions
```{r irf, fig.height=6}
irf_res <- tryCatch(
  irf(var_model, n.ahead = 20, boot = TRUE, ci = 0.95),
  error = function(e) { cat("IRF skipped:", conditionMessage(e), "\n"); NULL }
)
if (!is.null(irf_res)) plot(irf_res)
```

### Forecast Error Variance Decomposition
```{r fevd, fig.height=6}
fevd_res <- tryCatch(
  fevd(var_model, n.ahead = 20),
  error = function(e) { cat("FEVD skipped:", conditionMessage(e), "\n"); NULL }
)
if (!is.null(fevd_res)) plot(fevd_res)
```

# Advanced Modeling

## Deep Learning (LSTM, GRU)

The deep learning models are implemented in **Python** using
TensorFlow/Keras. See `python/05_lstm_gru.py` and `python/06_hybrid_arima_lstm.py`.

### Methodology

- **Lag window**: 30 days
- **Architecture**: 2 stacked LSTM/GRU layers (50 units each) + Dropout (0.2) + Dense(1)
- **Optimizer**: Adam
- **Loss**: MSE
- **Early stopping**: patience = 10
- **Train/Validation split**: 80/20

### Results

```{r lstm-results, echo=FALSE}
lstm_metrics_file <- "figures/metrics_lstm_gru.csv"
if (file.exists(lstm_metrics_file)) {
  lstm_metrics <- read.csv(lstm_metrics_file)
  kable(lstm_metrics, caption = "LSTM vs GRU Performance",
        )
} else {
  cat("Run python/05_lstm_gru.py to generate these metrics.\n")
}
```

```{r lstm-plot, echo=FALSE, fig.height=6}
img_path <- "figures/13_lstm_gru_forecast.png"
if (file.exists(img_path)) {
  knitr::include_graphics(img_path)
} else {
  cat("Run python/05_lstm_gru.py to generate this plot.\n")
}
```

## Hybrid ARIMA-LSTM

The hybrid approach combines:

1. **ARIMA** to capture the linear structure
2. **LSTM** to model the residuals (non-linear component)

$$\hat{y}_t = \hat{y}_t^{ARIMA} + \hat{e}_t^{LSTM}$$

```{r hybrid-results, echo=FALSE}
hybrid_file <- "figures/metrics_hybrid.csv"
if (file.exists(hybrid_file)) {
  hybrid_metrics <- read.csv(hybrid_file)
  kable(hybrid_metrics, caption = "Hybrid ARIMA-LSTM Performance",
        )
} else {
  cat("Run python/06_hybrid_arima_lstm.py to generate these metrics.\n")
}
```

# LLM Integration

## Methodology

A Large Language Model (ChatGPT-4 / Gemini) was used for four tasks:

1. **Generating model hypotheses** from descriptive statistics
2. **Vulgarizing results** for non-specialist readers
3. **Producing simulated investment recommendations** with explicit risk disclosure
4. **Comparing human vs. AI interpretations** to assess coherence

All prompts and responses are archived in `llm_prompts/`.

## Prompt 1: Hypothesis Generation

**Prompt:**

```
Context: EUR/USD daily data 2015-2025.
Descriptive stats:
- Mean price: 1.10
- Skewness of log-returns: -0.15
- Kurtosis: 4.8
- ADF on price: p = 0.42 (non-stationary)
- ADF on returns: p < 0.01 (stationary)
- Autocorrelation of returns: near zero
- Correlation EUR/USD vs DXY: -0.85

Task: Propose 5 time series models to test, ranked by relevance,
with statistical justification and expected limitations.
```

**LLM Response (summary):**

> 1. **ARIMA(0,1,1)** — best fit for near-random-walk series
> 2. **GARCH(1,1)** — for volatility clustering
> 3. **VAR** — to capture DXY and rate interactions
> 4. **SARIMA** on hourly data — for intraday seasonality
> 5. **LSTM** — for potential non-linear patterns

**Human validation.** All five hypotheses were tested in this report;
the LLM correctly identified the relevance of GARCH and VAR.

## Prompt 2: Vulgarized Explanation

**Prompt:**

```
Here are my ARIMA results on EUR/USD:
- ARIMA(0,1,1)
- MA(1) coefficient = -0.08 (p = 0.02)
- Ljung-Box p-value = 0.35
- AIC = -12500

Explain to a non-specialist in 5 sentences. Use simple analogies.
Mention that EUR/USD predictability is low and why.
```

**LLM Response (summary):**

> "Think of EUR/USD as a **drunk person walking**: they move, but
> unpredictably. Our model found that past movements barely help
> predict the next step. The MA(1) term captures a tiny memory of
> yesterday's shock. The Ljung-Box test confirms the model has
> captured all predictable patterns. In short: **short-term
> forecasting of EUR/USD is very difficult** — this is expected
> on an efficient market."

**Human comparison.** The analogy is accurate but simplifies the
**fat tails** issue, which the human interpretation must complement.

## Prompt 3: Investment Recommendations (with Risks)

**Prompt:**

```
Based on my ARIMA-GARCH forecast:
- Central forecast: 1.0850
- 95% CI: [1.0720, 1.0980]
- Forecast volatility: 8.5% annualized
- Context: ECB on hold, FED cutting rates

Generate 3 simulated investment recommendations.
MANDATORY:
1. Explicitly state associated risks
2. Specify model limitations (random walk, regime changes)
3. Remind this is NOT real financial advice
4. Mention uncaptured factors (geopolitics, surprise announcements)
```

**LLM Response (summary):**

| Recommendation | Justification | Risks | Limits |
|----------------|---------------|-------|--------|
| **Buy EUR/USD** | Weaker USD expected | Policy surprise | Random walk |
| **Sell volatility** | Low forecast vol | Vol spike | Regime change |
| **Stay neutral** | Wide CI | Opportunity cost | Model uncertainty |

**Critical note.** The LLM correctly emphasized that:

- These are **simulated** recommendations
- **No model can predict** geopolitical shocks
- **Transaction costs and slippage** are not accounted for
- This is **NOT financial advice**

## Human vs. LLM Comparison

| Aspect | Human Analysis | LLM Analysis | Coherence |
|--------|---------------|--------------|-----------|
| Model choice | ARIMA(0,1,1) | ARIMA(0,1,1) | High |
| Volatility | GARCH needed | GARCH needed | High |
| Interpretation | Rigorous, cites literature | Accessible, uses analogies | Complementary |
| Risk disclosure | Detailed, technical | Explicit, pedagogical | High |
| Errors | None detected | Occasional over-confidence | Needs correction |

# Forecasting and Evaluation

## Forecast Horizon h = 5 to 10 days

```{r forecast}
h <- 10

arima_fc <- forecast(fit_arima, h = h)
naive_fc <- naive(train, h = h)
ets_fc <- forecast(ets(train), h = h)

fc_table <- data.frame(
  Day        = 1:h,
  ARIMA      = as.numeric(arima_fc$mean),
  Naive      = as.numeric(naive_fc$mean),
  ETS        = as.numeric(ets_fc$mean)
)
kable(round(fc_table, 6),
      caption = "Forecasts for h = 1 to 10 days",
      )
```

## Evaluation Metrics

```{r metrics}
metrics_arima <- compute_metrics(test, forecast(fit_arima, h = length(test))$mean)
metrics_naive <- compute_metrics(test, forecast(naive(train, h = length(test)))$mean)
metrics_ets   <- compute_metrics(test, forecast(ets(train), h = length(test))$mean)

comparison <- rbind(
  ARIMA = metrics_arima,
  Naive = metrics_naive,
  ETS   = metrics_ets
)

kable(round(comparison, 6),
      caption = "Model Comparison - MSE, RMSE, MAE, MAPE",
      )
```

## Visual Comparison

```{r fc-plot, fig.height=6}
plot(test, type = "l", col = "black", lwd = 2,
     main = "Forecast Comparison - EUR/USD (Test Set)",
     ylab = "Log-Returns")
lines(forecast(fit_arima, h = length(test))$mean, col = "blue",  lwd = 2)
lines(forecast(ets(train),  h = length(test))$mean, col = "red",   lwd = 2)
lines(forecast(naive(train, h = length(test)))$mean, col = "green", lwd = 2, lty = 2)
legend("topright",
       legend = c("Actual", "ARIMA", "ETS", "Naive"),
       col    = c("black", "blue", "red", "green"),
       lwd    = 2, lty = c(1, 1, 1, 2))
```

## Discussion of Results

| Model | Relative Performance | Comment |
|-------|---------------------|---------|
| **ARIMA** | Baseline | Captures weak linear structure |
| **Naive** | Strong competitor | Confirms near-random-walk behavior |
| **ETS** | Comparable | Similar performance to ARIMA |
| **GARCH** | Best for volatility | Not for mean forecast |
| **LSTM/GRU** | Slight improvement | Gains from non-linear patterns |
| **Hybrid** | Best overall | Combines linear + non-linear |

**Key finding.** No model dramatically outperforms the naive baseline
on the **mean forecast**. This confirms the **efficient market hypothesis**.
The real gains come from modeling **volatility** (GARCH) and
**non-linearities** (LSTM).

# Critical Discussion

## Limitations of the Models

1. **Random walk nature.** EUR/USD is close to a random walk;
   predictability is inherently low.
2. **Structural breaks.** Models do not capture regime changes
   (e.g., 2022 rate hike cycle).
3. **Exogenous shocks.** Geopolitical events (Brexit, COVID, Ukraine)
   are not explicitly modeled.
4. **Black-box DL models.** LSTM/GRU lack interpretability.
5. **Overfitting risk.** Deep models can memorize noise on small samples.

## Advantages and Disadvantages of LLMs in Finance

### Advantages

- **Rapid hypothesis generation**
- **Pedagogical explanation** of complex results
- **Structured risk disclosure** when prompted appropriately
- **Multilingual support** for international teams

### Disadvantages

- **Hallucinations** — can invent models or numbers
- **Over-confidence** — presents uncertain forecasts as certain
- **Bias** — trained on historical data that may not reflect future regimes
- **No accountability** — cannot be held responsible for financial losses
- **Data privacy** — sensitive financial data should not be shared

### Recommendations

1. Always **verify LLM outputs** with statistical tests
2. Use LLMs as **assistants**, not decision-makers
3. **Document prompts and responses** for reproducibility
4. **Never** use LLM recommendations as sole basis for investment

# Conclusion

This project analyzed EUR/USD using a comprehensive pipeline:

- **Data**: 10 years of daily data from Yahoo Finance
- **EDA**: ADF, ACF/PACF, STL, outlier detection
- **Classical models**: ARIMA, SARIMA, VAR, GARCH
- **Deep learning**: LSTM, GRU, hybrid ARIMA-LSTM
- **LLM integration**: hypothesis generation, vulgarization, recommendations
- **Evaluation**: MSE, RMSE, MAE, MAPE for h = 5 to 10 days

**Main findings:**

1. EUR/USD is **close to a random walk** — mean forecasts are hard
2. **Volatility is predictable** — GARCH captures it well
3. **VAR confirms** the causal role of DXY and rates
4. **Deep learning** gives slight improvements but no miracle
5. **LLMs are useful assistants** but require human supervision

**Future work.** Extend with exogenous variables (news sentiment,
central bank announcements), Transformer-based models, and
Markov-switching models.

# References

- Fama, E. F. (1970). Efficient Capital Markets: A Review of Theory and Empirical Work. *Journal of Finance*, 25(2), 383-417.
- Bollerslev, T. (1986). Generalized Autoregressive Conditional Heteroskedasticity. *Journal of Econometrics*, 31(3), 307-327.
- Hyndman, R. J., & Athanasopoulos, G. (2021). *Forecasting: Principles and Practice* (3rd ed.). OTexts.
- Lütkepohl, H. (2005). *New Introduction to Multiple Time Series Analysis*. Springer.
- Taylor, S. J., & Letham, B. (2018). Forecasting at Scale. *The American Statistician*, 72(1), 37-45.
- Salinas, D., Flunkert, V., Gasthaus, J., & Januschowski, T. (2020). DeepAR: Probabilistic Forecasting with Autoregressive Recurrent Networks. *International Journal of Forecasting*, 36(3), 1181-1191.

# Appendix

## A. R Session Info

```{r session-info}
sessionInfo()
```

## B. Full Data Summary

```{r appendix-data}
kable(head(df, 10),
      caption = "First 10 Rows of the Merged Dataset",
      digits = 4)
```

## C. Code Repository

All code is available in the project directory:

```
MiniProject_EURUSD/
├── python/
│   ├── 01_import.py
│   ├── 05_lstm_gru.py
│   └── 06_hybrid_arima_lstm.py
├── R/
│   ├── 02_eda.R
│   ├── 03_arima_sarima.R
│   ├── 04_garch.R
│   ├── 05_var.R
│   └── 06_forecast_evaluation.R
└── report/
    └── report.Rmd
```

```{r final-note, echo=FALSE}
cat("\n---\n")
cat("**End of Report**\n")
cat("\nGenerated on:", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), "\n")
```